---
title: 'SOA Predictive Analytics Exam, Module 8, Section 1'

---
If you are using R 3.6.0 or later, the following command will ensure that the random number generator being used is the same as in prior versions of R. Because these modules were developed using R 3.5.0, output that depends on random numbers used the older generator. Running this code will have your output match that shown in the modules.

```{r}
# Run this chunk if using R 3.6.0 or later.
RNGkind(sample.kind = "Rounding")
```

If you are using R 4.0.0 or later, the following command will ensure that read.csv() interprets variables whose values are characters as factor variables. This was the default behavior in prior versions of R. All code is written assuming such variables are factor variables.

```{r}
# Run this chunk if using R 4.0.0 or later. You may get what looks like an error message. It can be ignored.
options(stringsAsFactors = TRUE)
```

Run CHUNK 1 to load two libraries and the diamonds dataset as part of Exercise 8.1.1.

```{r}
# CHUNK 1
library(ggplot2)
library(gridExtra)
diamonds <- read.csv(file = "diamonds.csv")
head(diamonds)
summary(diamonds)
```

Run CHUNK 2 as part of Exercise 8.1.1.

```{r}
# CHUNK 2

# Hints
# 1. We find the center of the data, which is about (6.5, 4)
# 2. The line of maximum variance has a slope of about 1.75
# 3. Draw the first principal component (red)
# 4. Second principal component should be orthogonal (perpendicular) to the first principal component
# 5. Draw the second principal component (purple)

p1 <- ggplot(diamonds, aes(x = x, y = z)) +
  geom_point() +
  xlim(3, 10) +
  ylim(0, 7.5)
p1
```

Run CHUNK 3 to see the solution.

```{r}
# CHUNK 3
# 1. We find the center of the data, which is about (6.5, 4)
# 2. The line of maximum variance has a slope of about 1.75
# 3. Draw the first principal component (red)
# 4. Second principal component should be orthogonal (perpendicular) to the first principal component
# 5. Draw the second principal component (purple)

p2 <- p1 + geom_segment(aes(x = 6.5, y = 4, xend = 10, yend = 6),
  color = "red",
  arrow = arrow(length = unit(0.1, "npc"))
) +
  ggtitle("1st Principal Component")
p3 <- p2 + geom_segment(aes(x = 6.5, y = 4, xend = 6.1, yend = 4.7),
  color = "purple",
  arrow = arrow(length = unit(0.1, "npc"))
) +
  ggtitle("1st & 2nd Principal Components")
grid.arrange(p2, p3, ncol = 2)
```

Run CHUNK 4 to see a non-orthogonal version.
  
```{r}
# CHUNK 4
p4 <- p1 + geom_segment(aes(x = 6.5, y = 4, xend = 10, yend = 6),
  colour = "red",
  arrow = arrow(length = unit(0.1, "npc"))
) +
  geom_segment(aes(x = 6.5, y = 4, xend = 10, yend = 6.5),
    colour = "purple",
    arrow = arrow(length = unit(0.1, "npc"))
  ) +
  ggtitle("1st & 2nd principal Components")
p4
```

CHUNK 5 provides some information about variances.

```{r}
# CHUNK 5
# Subset the data to be consistent with previous plot
diamonds.2d.pca <- subset(diamonds[, c(8, 10)], x >= 3 & x <= 10 & z > 0 & z <= 7.5)
# Calculate PCA for diamond length and depth
# Note we center (subtract the sample mean) and scale (divide by the sample standard deviation) the data be setting center = TRUE and scale. = TRUE
# The function prcomp comes with base R, so no package needs to be loaded
d.pca.2d <- prcomp(diamonds.2d.pca, center = TRUE, scale. = TRUE)
summary(d.pca.2d)
```

Run CHUNK 6 to add more variables.

```{r}
# CHUNK 6
# Add more variables to the data
diamonds.5d.pca <- subset(diamonds[, c(5, 7:10)], x >= 3 & x <= 10 & z > 0 & z <= 7.5)
# Calculate PCA
d.pca.5d <- prcomp(diamonds.5d.pca, center = TRUE, scale. = TRUE)
summary(d.pca.5d)
```  

Run CHUNK 7 to display the loadings.

```{r}
# CHUNK 7
d.pca.5d$rotation
```    

Run CHUNK 7A to calculate the first principal component and display the first six values.

```{r}
# CHUNK 7A
# First do it manually
loadings <- d.pca.5d$rotation[, 1] # Extract the loadings for PC1
loadings
diamonds.standardized <- scale(diamonds.5d.pca) # standardize the data
PC1 <- diamonds.standardized %*% loadings
head(PC1)

# Then use R
head(d.pca.5d$x[, 1]) # The variable x holds the PCs.
```

Run CHUNK 7B to manually calculate the variances and standard deviations.

```{r}
# CHUNK 7B
Variances <- diag(var(d.pca.5d$x))
Variances
sqrt(Variances)
summary(d.pca.5d) # Verifies that the manual standard deviations match the output.
```

Run CHUNK 8 to compare centering versus scaling.

```{r}
# CHUNK 8
# Sample diamonds data due to speed of plotting
set.seed(1000)
diamonds.sample <- diamonds[sample(1:nrow(diamonds), 5000), ]
diamonds.3d.sample <- subset(diamonds.sample[, c(8:10)], x >= 3 & x <= 10 & z > 0 & z <= 7.5)
diamonds.5d.sample <- subset(diamonds.sample[, c(5, 7:10)], x >= 3 & x <= 10 & z > 0 & z <= 7.5)

# Compute PCA for the two samples, one scaled and one without
d.3d.scale <- prcomp(diamonds.3d.sample, center = TRUE, scale. = TRUE)
d.3d.noscale <- prcomp(diamonds.3d.sample, center = TRUE, scale. = FALSE)
d.5d.scale <- prcomp(diamonds.5d.sample, center = TRUE, scale. = TRUE)
d.5d.noscale <- prcomp(diamonds.5d.sample, center = TRUE, scale. = FALSE)

# Plot the four graphs
par(mfrow = c(1, 2))
p1 <- biplot(d.3d.scale, scale = 0) # main = "3D Scaled"
p2 <- biplot(d.3d.noscale, scale = 0) # main = "3D No Scale"
p3 <- biplot(d.5d.scale, scale = 0) # main = "5D Scaled"
p4 <- biplot(d.5d.noscale, scale = 0) # main = "5D No Scale"
```
  
Run CHUNK 9 to see more about the three variable PCA.

```{r}
# CHUNK 9
# What do you see?
summary(d.3d.scale)
d.3d.scale$rotation
summary(d.3d.noscale)
d.3d.noscale$rotation
```

Run CHUNK 10 to see the results for the five variable problem.

```{r}
# CHUNK 10
# What do we see here?
# The output of unscaled PCA doesn't make much sense as price dominated all the variance
summary(d.5d.scale)
d.5d.scale$rotation
summary(d.5d.noscale)
d.5d.noscale$rotation
```
 
 Use CHUNKS 11-13 to further interpret the dimensionality of the diamonds data.

```{r}
# CHUNK 11
# Continue with the diamonds.sample data (5000 obs). Binarize all categorical variables
```

```{r}
# CHUNK 12
# Conduct PCA using all variables except for price and make a simple plot. You might like to try the biplot() and screeplot() functions.
# Do you know why we leave the price out?
```

```{r}
# CHUNK 13
# Combine the plot with other PCA outputs (summary, rotations etc). What are your observations?
# Note - interpretation of the principal components is based on finding which variables are most strongly correlated with each component.
# In other words, we need to find which numbers are large among all the coefficients.
```

Examine CHUNKS 14-16 to see a possible solution.

```{r}
# CHUNK 14
library(caret)

# Binarize all categorical variables
# dummyVars is not compatible with factors
diamonds$color <- as.character(diamonds$color)
diamonds$cut <- as.character(diamonds$cut)
diamonds$clarity <- as.character(diamonds$clarity)
# Binarize variables
binarizer <- caret::dummyVars("~ cut + clarity + color", data = diamonds, fullRank = F)
diamonds_spread <- cbind(diamonds, data.frame(predict(binarizer, newdata = diamonds)))

# Remove price
diamonds_spread <- diamonds_spread[, setdiff(colnames(diamonds_spread), c("price", "color", "cut", "clarity"))]

# Quick view
head(diamonds_spread)
```


```{r}
# CHUNK 15
# Conduct PCA using all variables except for price and make a simple plot
# Do you know why we leave the price out?
# PCA is an unsupervised technique. If we are to use the PCs later on in a model, the target cannot be included in the analysis.
d.pca <- prcomp(diamonds_spread, center = TRUE, scale. = TRUE)
par(mfrow = c(1, 2))
biplot(d.pca)
screeplot(d.pca, npcs = 26, type = "lines")
```


```{r}
# CHUNK 16
# Combine the plot with other PCA outputs. What are some of your observations?
# From the scree plot it appears the first 19 PCs explain most of the variance. From the output in the next chunk, it turns out that they explain 98% of the variance. That is a high number and perhaps fewer might be used, though it not clear where the break is.
# We want to find the factors driving each PC. One way we can do this is to rank or order coefficients within each PC and make an observation, e.g. try
# abs.load <- abs(d.pca$rotation)
# ordered.abs.load <- rev(order(abs.load[,1]))
# head(abs.load[ordered.abs.load,1])

summary(d.pca)
d.pca$rotation
```

In this example, the first principal component is a measure of "size" where x, y, z and carat are the dominating variables. These all have the same sign, so increases (or decreases) have the same effect. Pay attention to the positive values for the highest quality categories of cut (Ideal), clarity (IF) and color (D). They are negatively correlated with the size measures.
